Volatility Metric to Detect Anomalies in Source Code Repositories

### 2020-04-19 paper/article.tex
more

\usepackage[utf8]{inputenc}
\usepackage{textcomp}
\usepackage[inline]{enumitem}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{verbatimbox}
\usepackage{interval}
\usepackage{hyperref}
\usepackage{minted}
  \setminted{fontsize=\footnotesize}
  \setminted{breaklines}
  \usemintedstyle{bw}
\newcommand{\code}[1]{\texttt{#1}}
\newenvironment{nicetable}
  {\setlength{\parindent}{0em}\medskip\small}
  {\medskip}
\input{total}
\begin{document}
\begin{abstract}
A new metric was introduced to calculate the distance
between actively modified files in a source code repository
and the files, which are rarely modified and may be considered
as ``dead'' or ``abandoned.'' It was empirically demonstrated that larger repositories
have larger values of the introduced metric.
The metric may be used for earlier detection of code maintenance anomalies
and helping software developers make the decision of splitting the repository
into smaller ones in order to prevent maintainability issues.
\end{abstract}
\maketitle

@@

Git logs provide information about every single change made by every software
developer during the entire course of the project. Using this information
it's possible to measure which files are being modified frequently. On the
other hand, it's also possible to spot files that are rarely modified and may
be considered as ``dead'' or ``abandoned'' code, which may be considered as a threat
to the maintainability of the entire project.

@@

The variance $\text{Var}(g)$ is the volatility of the source code. The smaller
the volatility the more cohesive is the repository and the smaller
the amount of the dead code inside it.

@@

The Figure~\ref{fig:1} demonstrates the relationship between
the number of files in the repository~($M_1$) and its volatility~($V$). Both
axixes of the graph have logarithmic scales, for the sake of visual
understandability: the difference between minimum and maximum values
of volatility is logarithmically large. It is visually obvious that
repositories with larger number of files tend to have higher values
of the volatility metric.

\begin{figure}[h]
  \input{files.tex}
  \caption{The relationship between the number of files in the repository and its volatility}
  \label{fig:1}
\end{figure}

@@

\begin{figure}[h]
  \input{bytes.tex}
  \caption{The relationship between the size of the Git repository in bytes and its volatility}
  \label{fig:2}
\end{figure}

@@

It was demonstrated that the size of the source code is an important
factor negatively affecting maintainability and other quality characteristicts.
For example, \citet{heitlager2007} suggested to consider ``volume'' (the amount of code
in the repository, the number of lines per file, the number of methods in a class,
and so on) as a primary contributor to an aggregate maintainability index
of a software because ``it is fairly intuitive that the total size of a system should
feature heavily in any measure of maintainability.''

>>>>

\usepackage[utf8]{inputenc}
\usepackage{textcomp}
\usepackage[inline]{enumitem}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{verbatimbox}
\usepackage{interval}
\usepackage{hyperref}
\usepackage{minted}
  \setminted{fontsize=\footnotesize}
  \setminted{breaklines}
  \usemintedstyle{bw}
\newcommand{\code}[1]{\texttt{#1}}
\newenvironment{nicetable}
  {\setlength{\parindent}{0em}\medskip\small}
  {\medskip}
\input{total}
\begin{document}
\begin{abstract}
A new metric was introduced to calculate the distance
between actively modified files in a source code repository
and the files, which are rarely modified and may be considered
abandoned or even dead. It was empirically demonstrated that larger repositories
have larger values of the introduced metric.
The metric may be used for earlier detection of code maintenance anomalies
and helping software developers make the decision of splitting the repository
into smaller ones in order to prevent maintainability issues.
\end{abstract}
\maketitle

@@

Git logs provide information about every single change made by every software
developer during the entire course of the project. Using this information
it's possible to measure which files are being modified frequently. On the
other hand, it's also possible to spot files that are rarely modified and may
be considered as abandoned code, which may be considered as a threat
to the maintainability of the entire project.

@@

The variance $\text{Var}(g)$ is the volatility of the source code. The smaller
the volatility the more cohesive is the repository and the smaller
the amount of the abandonded code inside it.

@@

The Figure~\ref{fig:1} demonstrates the relationship between
the number of files in the repository~($M_1$) and its volatility~($V$). Both
axixes of the graph have logarithmic scales, for the sake of visual
understandability: the difference between the minimum and the maximum values
of the volatility is logarithmically large. It is visually obvious that
repositories with larger number of files tend to have higher values
of the volatility metric.

\begin{figure}[h]
  \input{files-64.tex}
  \caption{The relationship between the number of files in the repository and its volatility ($Z$ is set to 64)}
  \label{fig:1}
\end{figure}

@@

\begin{figure}[h]
  \input{bytes-64.tex}
  \caption{The relationship between the size of the Git repository in bytes and its volatility ($Z$ is set to 64)}
  \label{fig:2}
\end{figure}

The Figure~\ref{fig:1} demonstrates the same relationship as the Figure~\ref{fig:1},
but the parameter of the volatility formula is set to 32 instead of 64. The
trend of the graph didn't change much.

\begin{figure}[h]
  \input{files-32.tex}
  \caption{The relationship between the number of files in the repository and its volatility ($Z$ is set to 32)}
  \label{fig:3}
\end{figure}

The Figure~\ref{fig:1} demonstrates the same relationship as the Figure~\ref{fig:1},
but this time the parameter of the volatility formula is increased to 128
instead of the original value of 64. The trend of the graph didn't change much.

\begin{figure}[h]
  \input{files-128.tex}
  \caption{The relationship between the number of files in the repository and its volatility ($Z$ is set to 128)}
  \label{fig:4}
\end{figure}

@@

It was demonstrated that the size of the source code is an important
factor negatively affecting maintainability and other quality characteristicts.
For example, \citet{heitlager2007} suggested to consider ``volume''
(the number of lines per file, the number of methods in a class,
and so on) as a primary contributor to an aggregate maintainability index
of a software because ``it is fairly intuitive that the total size of a system should
feature heavily in any measure of maintainability.''

However, the discussion of whether larger repositories are a better
practice than smaller ones is still open. Google, for example, advocates
for larger ones, known as ``monolithic,'' which tend to keep many modules
and components in one Git space, simplifying dependencies management~\citep{jaspan2018}.
As was demonstrated in our research, larger repositories will tend to
have maintenance anomalies: some parts of source code will be changed much
less frequently than other parts. Moreover, the distance between ``popular''
parts and ``abandoned'' ones will grow when the size of the repository grows.

These maintenance anomalies will lead to maintability and quality problems
mostly because it will be difficult for programmers to quickly switch between
different contexts inside one repository: one set of files is being modified
every day, while another one stays untouched for months. A programmer will
try to avoid making changes to the part that is abandonded because it won't
look like code that is expecting changes.

Also, maintenance anomalies is a sign of differences in customer demand
for the source code components. The module with smaller frequency of changes
must be released less frequently than the module that is being edited every day.
Not paying attention to the growing volatility of the repository means
ignoring anomalies and, because of that, redundant releasing of the code
that doesn't require new releases. Moreover, releasing larger software modules
usually means longer build cycles, which are the key failure factor of
continuous delivery, as explained by~\cite{humble2010}. Getting rid of
maintenance anomalies may help reduce build cycles.
